\begin{flushleft}
I would like to thank Prof. Dr. Michael Boutros for providing the opportunity to work on this ambitious thesis project. I am grateful for the regular feedback and all the support received during my work.
\bigbreak

I appreciated the support from Dr. Tianzuo Zhan who supervised me early during my work in the laboratory. Without his support and encouragement to independently pursue the establishment of organoid models, this project would not have developed into its current form.  
\bigbreak

I am truly grateful to Dr. Johannes Betge who was a wonderful colleague and scientific partner. He joined the laboratory at a time when the isolation and image-based profiling was still at a  prototype stage. His support enabled the establishment and profiling of patient derived organoid models. With him, I found a great colleague to pursue this project. We spent many very long evenings in the cell culture laboratory isolating and expanding organoid models. Without his support, the presented work would not have been possible. 
\bigbreak

I would like to thank Dr. Jan Sauer for his work on image-analysis of 3D organoid cultures. My thanks also go to Dr. Benedikt Rauscher for thoughtful advice throughout this project and an initial analysis of the patient derived organoid transcript expression data.  
\bigbreak

I am especially thankful to Dr. Kim Boonekamp, who supported me during the completion of this thesis as well as in the discussion of mouse organoid profiling results with very thoughtful comments. Without her scientific and personal support, this thesis would not have been shaped into this current form.
\bigbreak

My sincere appreciation goes to the engineering team of the Signaling and Functional Genomics group, especially Thilo Miersch whose expertise in high-throughput liquid handling was instrumental to making the image-based profiling method work for organoid models at scale. 
\bigbreak

Lastly, I extend my heartfelt thanks to all members of the Signaling and Functional Genomics group. The laboratory has been an ideal place for my scientific training and first independent steps. It was a pleasure to work with you. 
\bigbreak

I am dedicating this thesis to my family.
\end{flushleft}